How do changes in ex vary between countries?

### 2022-03-12 draft_working_paper.Rmd
outstanding commits showing - so ticking stage and pushing!

---
title: "How do changes in ex vary between countries?"
author: "Jon Minton, Lucinda Hiam"
date: "22/02/2022"
output:
  html_document: default
  word_document: default
---

@@

```{r, echo=FALSE}
# load packages

pacman::p_load(knitr, here, tidyverse, flextable) # using flextable instead of gt. It seems to be better 
# developed, more commonly used, and has more clearly labelled function names.

# load data

hmd_lt <- read_rds(here("data", "lifetables.rds"))

@@

```{r, echo = FALSE}

countries_of_interest <- c(
  "GBRTENW",
  "GBR_SCO",
  "GBR_UK",
  "FRATNP",
  "ESP",
  "ITA",
  "DEUTNP",
  "DEUTE", 
  "DEUTW"
)

hmd_ex_selected_countries <- 
  hmd_lt %>% 
    filter(code %in% countries_of_interest) %>% 
    filter(Age %in% c(0, 65)) %>% 
    select(code, Year, Age, sex, ex) %>% 
    rename(x = Age, year = Year)

hmd_ex_selected_countries

# Code for producing synthetic Germany

source(here("R", "make_synthetic_germany_functions.R"))

series_east_f0 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTE") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 0) %>% 
  filter(sex == "female") %>% 
  arrange(year) %>% 
  pull(ex)

series_west_f0 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTW") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 0) %>% 
  filter(sex == "female") %>% 
  arrange(year) %>% 
  pull(ex)

tmp_df_f0 <- tibble(
  code = "DEUTSYNTH",
  year = 1980:2017,
  x = 0,
  sex = "female", 
  ex = make_synthetic_population(series_east_f0, series_west_f0, p_east = 0.2)
)

###
series_east_f65 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTE") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  filter(sex == "female") %>% 
  arrange(year) %>% 
  pull(ex)

series_west_f65 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTW") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  filter(sex == "female") %>% 
  arrange(year) %>% 
  pull(ex)

tmp_df_f65 <- tibble(
  code = "DEUTSYNTH",
  year = 1980:2017,
  x = 65,
  sex = "female", 
  ex = make_synthetic_population(series_east_f65, series_west_f65, p_east = 0.2)
)
##

series_east_m0 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTE") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 0) %>% 
  filter(sex == "male") %>% 
  arrange(year) %>% 
  pull(ex)

series_west_m0 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTW") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 0) %>% 
  filter(sex == "male") %>% 
  arrange(year) %>% 
  pull(ex)

tmp_df_m0 <- tibble(
  code = "DEUTSYNTH",
  year = 1980:2017,
  x = 0,
  sex = "male", 
  ex = make_synthetic_population(series_east_m0, series_west_m0, p_east = 0.2)
)

series_east_m65 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTE") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  filter(sex == "male") %>% 
  arrange(year) %>% 
  pull(ex)

series_west_m65 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTW") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  filter(sex == "male") %>% 
  arrange(year) %>% 
  pull(ex)

tmp_df_m65 <- tibble(
  code = "DEUTSYNTH",
  year = 1980:2017,
  x = 65,
  sex = "male", 
  ex = make_synthetic_population(series_east_m65, series_west_m65, p_east = 0.2)
)

hmd_ex_selected_countries_with_synth <- bind_rows(
  hmd_ex_selected_countries,
  tmp_df_m0,
  tmp_df_m65,
  tmp_df_f0,
  tmp_df_f65
)

# Labels for codes

country_code_lookup <- 
  tribble(
    ~code, ~country,
    "DEUTNP", "Germany",
    "DEUTE", "East Germany",
    "DEUTW", "West Germany",
    "ESP", "Spain",
    "FRATNP", "France", 
    "ITA", "Italy",
    "GBRTENW", "England & Wales",
    "GBR_SCO", "Scotland",
    "DEUTSYNTH", "Synthetic Germany"
  )

@@

```{r, echo = FALSE, fig.cap = "Life Expectancy at birth"}

hmd_ex_selected_countries_with_synth %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy"))) %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 0) %>% 
  ggplot(aes(x = year, y = ex, group = sex, colour = sex, linetype = sex)) + 
  geom_line() +
  facet_wrap(~country) +
  labs(
    x = "Year",
    y = "Life Expectancy at birth",
    title = "Life expectancies at birth for selected nations",
    subtitle = "1980 to 2020 or latest available year",
    caption = "Source: Human Mortality Database"
  )

```{r, echo = FALSE, fig.cap = "Life Expectancy at age 65"}
hmd_ex_selected_countries_with_synth %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy"))) %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  ggplot(aes(x = year, y = ex, group = sex, colour = sex, linetype = sex)) + 
  geom_line() +
  facet_wrap(~country) +
  labs(
    x = "Year",
    y = "Life Expectancy at age 65",
    title = "Life expectancies at age 65 for selected nations",
    subtitle = "1980 to 2020 or latest available year",
    caption = "Source: Human Mortality Database"
  )

- [ ] Some insightful comment about what's going on.

@@

```{r, echo = FALSE, warning=FALSE, message = FALSE, fig.height=8, fig.width=8}
change_in_ex_selected_countries %>% 
  filter(x == 0) %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy"))) %>% 
  filter(between(year, 1980, 2020)) %>% 
  ggplot(aes(x = year, y = delta_ex, colour=sex)) + 
  geom_point(alpha = 0.2) + 
  stat_smooth() + 
  facet_grid(country~sex) + 
  geom_hline(yintercept = 0) + 
  theme(strip.text.y = element_text(angle = 0)) + 
  labs(
    x = "Year",
    y = "Year-on-year change in life expectancy",
    title = "Annual change in life expectancy at birth, selected countries",
    subtitle = "Blue line: nonlinear smoother",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )

@@

```{r, echo = FALSE, warning=FALSE, message = FALSE, fig.height=8, fig.width=8}
change_in_ex_selected_countries %>% 
  filter(x == 65) %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy"))) %>% 
  filter(between(year, 1980, 2020)) %>% 
  ggplot(aes(x = year, y = delta_ex, colour=sex)) + 
  geom_point(alpha = 0.2) + 
  stat_smooth() + 
  facet_grid(country~sex) + 
  geom_hline(yintercept = 0) + 
  theme(strip.text.y = element_text(angle = 0)) + 
  labs(
    x = "Year",
    y = "Year-on-year change in life expectancy",
    title = "Annual change in life expectancy at age 65, selected countries",
    subtitle = "Blue line: nonlinear smoother",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )

@@

```{r, echo = FALSE, warning = FALSE, message = FALSE}

hmd_ex_selected_countries_with_synth %>% 
    filter(year >= 1979) %>% 
    filter(!(code %in% c("DEUTE", "DEUTW", "DEUTNP"))) %>%  # Using only synthetic germany for longer time period
    group_by(code, x, sex) %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
    nest() %>% 
    mutate(
      rwd_model = map(data, ~lm(delta_ex ~ 1, data = . )),
      ar_model  = map(data, ~lm(delta_ex ~ lag(delta_ex), data = .))
    ) %>% 
    mutate(
      aic_rwd = map_dbl(rwd_model, AIC),
      aic_ar  = map_dbl(ar_model, AIC)
    ) %>% 
    mutate(
      diff_aic = aic_ar - aic_rwd
    )  %>% 
  # filter(diff_aic > 0)
  mutate(
    tidied_ar_model = map(ar_model, broom::tidy)
  ) %>% 
  unnest(tidied_ar_model) %>% 
  filter(term == "(Intercept)") %>% 
  arrange(desc(estimate)) %>%
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy"))) %>% 
  select(code, country, x, sex, estimate, std.error) %>% 
  mutate(display_label = glue::glue("{country}_{x}_{sex}")) %>% 
  ggplot(aes(estimate, fct_reorder(display_label, estimate))) + 
  geom_point() +
  expand_limits( x = 0) + 
  geom_vline(xintercept = 0) + 
  labs(
    x = "Average annual life expectancy change",
    y = "Country-sex combination",
    title = "Average annual improvements by country and sex",
    subtitle = "Range: 1980 to 2020 or nearest available years",
    caption = "Source: Human Mortality Database"
  ) + 
  facet_wrap(~x)

@@

### Average improvements by decade

```{r, echo=FALSE, tab.cap = "Source: HMD"}

hmd_ex_selected_countries_with_synth %>% 
    filter(year >= 1979) %>% 
    group_by(code, x, sex) %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>%
    mutate(decade = year - (year %% 10)) %>% 
    filter(!is.na(delta_ex)) %>% 
  group_by(code, x, sex, decade) %>% 
  summarise(
    mean_delta_ex = mean(delta_ex),
    sd_delta_ex   = sd(delta_ex)
    ) %>% 
  ungroup() %>% 
  mutate(cell_contents = glue::glue("{round(mean_delta_ex, 3)} ({round(sd_delta_ex, 3)})")
  ) %>% 
  filter(code == "GBRTENW") %>% 
  filter(x == 0) %>% 
  select(sex, decade, cell_contents) %>% 
  spread(decade, cell_contents) %>% 
#  write.csv("clipboard")
  flextable() %>%
  # Want to make the header row bold
  vline(j = 1, part = "all") %>%
  set_caption(caption = "Average improvement in life expectancy by decade")

```{r, cache=TRUE}

dta_for_decade_comparison <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 0) %>% 
      filter(sex == "female") %>% 
      filter(code == "GBRTENW") %>% 
      filter(year >= 1979) %>% 
      # ggplot(aes(x = year, y = ex)) +
      #   geom_line()
      arrange(year) %>% 
      mutate(delta_ex = ex - lag(ex)) %>% 
    #   ggplot(aes(x = year, y = delta_ex)) + 
    # geom_point() + geom_line() +
    # geom_hline(yintercept = 0)
      mutate(
        decade = year - (year %% 10)
      ) %>% 
  filter(!is.na(delta_ex))

model_null <- 
  lm(delta_ex ~ 1, data = dta_for_decade_comparison)

model_alt <- 
  lm(delta_ex ~ factor(decade), data= dta_for_decade_comparison)

summary(model_null)

summary(model_alt)

anova(model_null, model_alt)

# to get the value of interest do tmp[2,6]

# Generalising

tidy_dataset <- function(df){
  df %>% 
      filter(year >= 1979) %>% 
      arrange(year) %>% 
      mutate(delta_ex = ex - lag(ex)) %>% 
      mutate(
        decade = year - (year %% 10)
      ) %>% 
  filter(!is.na(delta_ex))
}

make_null_model <- function(df){
  lm(delta_ex ~ 1, data = df)
}

all_null_alt_models <-
  hmd_ex_selected_countries_with_synth %>%
    group_by(x, sex, code) %>% 
    nest() %>% 
    mutate(
      prep_data = map(data, tidy_dataset) 
    ) %>% 
    mutate(
      model_null = map(prep_data, ~lm(delta_ex ~ 1, data = .)),
      model_alt  = map(prep_data, ~lm(delta_ex ~ factor(decade), data = .))
  #    model_null = map(prep_data, make_null_model)
    ) %>% 
    mutate(
      anova_results = map2(model_null, model_alt, anova)
    ) %>% 
    mutate(
      anova_pval = map_dbl(anova_results, ~.[2,6])
    )

all_null_alt_models %>% 
  filter(anova_pval < 0.05) -> tmp

summary(tmp$model_alt[[1]])

summary(tmp$model_alt[[2]])

  # build null model
  # build alt model
  # do anova
  # get p value from anova

@@

NB: those with statistically significant 'breakpoints' in annual change in life expectancy are:
- Germany (need to re-download from GIT)

@@

  ex_ENW + labs(x = "Year", y = "Annual change in life expectancy", title = "Scotland", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
```

@@

  ex_ENW + labs(x = "Year", y = "Annual change in life expectancy", title = "France", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
```

@@

  ex_ENW + labs(x = "Year", y = "Annual change in life expectancy", title = "Spain", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
```

@@

  ex_ENW + labs(x = "Year", y = "Annual change in life expectancy", title = "Italy", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
```

@@

 #Grid
 ex_DEUTSYNTH <- hmd_ex_selected_countries_with_synth %>%
      filter(sex %in% c("female", "male")) %>%
      filter(code == "DEUTHSYNTH") %>%
    group_by(sex, x) %>% arrange(year) %>%
      mutate(delta_ex = ex - lag(ex)) %>%
    ungroup %>%
    filter(year >= 1979) %>%
    ggplot(aes(x = year, y = delta_ex)) + geom_point() +
   geom_line() +
   geom_hline(yintercept = 0) +
    stat_smooth(mapping = aes(colour = sex)) +
       facet_grid(x ~ sex)

  ex_ENW + labs(x = "Year", y = "Annual change in life expectancy", title = "Germany: Synthetic", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
```

@@

  ex_ENW + labs(x = "Year", y = "Annual change in life expectancy", title = "Germany: East", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
```

@@

  ex_ENW + labs(x = "Year", y = "Annual change in life expectancy", title = "Germany: West", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
```

>>>>

---
title: "How do changes in ex vary between countries?"
author: "Jon Minton, Lucinda Hiam"
date: "22/02/2022"
output: word_document
---

@@

```{r, echo=FALSE}
# load packages
pacman::p_load(knitr, here, tidyverse, flextable) # using flextable instead of gt. It seems to be better 
# developed, more commonly used, and has more clearly labelled function names.
# load data 
hmd_lt <- read_rds(here("data", "lifetables.rds"))
```

@@

```{r, echo = FALSE}
countries_of_interest <- c(
  "GBRTENW",
  "GBR_SCO",
  "GBR_UK",
  "FRATNP",
  "ESP",
  "ITA",
  "DEUTNP",
  "DEUTE", 
  "DEUTW"
)
hmd_ex_selected_countries <- 
  hmd_lt %>% 
    filter(code %in% countries_of_interest) %>% 
    filter(Age %in% c(0, 65)) %>% 
    select(code, Year, Age, sex, ex) %>% 
    rename(x = Age, year = Year)
hmd_ex_selected_countries
# Code for producing synthetic Germany
source(here("R", "make_synthetic_germany_functions.R"))
series_east_f0 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTE") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 0) %>% 
  filter(sex == "female") %>% 
  arrange(year) %>% 
  pull(ex)
series_west_f0 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTW") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 0) %>% 
  filter(sex == "female") %>% 
  arrange(year) %>% 
  pull(ex)
tmp_df_f0 <- tibble(
  code = "DEUTSYNTH",
  year = 1980:2017,
  x = 0,
  sex = "female", 
  ex = make_synthetic_population(series_east_f0, series_west_f0, p_east = 0.2)
)
###
series_east_f65 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTE") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  filter(sex == "female") %>% 
  arrange(year) %>% 
  pull(ex)
series_west_f65 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTW") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  filter(sex == "female") %>% 
  arrange(year) %>% 
  pull(ex)
tmp_df_f65 <- tibble(
  code = "DEUTSYNTH",
  year = 1980:2017,
  x = 65,
  sex = "female", 
  ex = make_synthetic_population(series_east_f65, series_west_f65, p_east = 0.2)
)
##
series_east_m0 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTE") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 0) %>% 
  filter(sex == "male") %>% 
  arrange(year) %>% 
  pull(ex)
series_west_m0 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTW") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 0) %>% 
  filter(sex == "male") %>% 
  arrange(year) %>% 
  pull(ex)
tmp_df_m0 <- tibble(
  code = "DEUTSYNTH",
  year = 1980:2017,
  x = 0,
  sex = "male", 
  ex = make_synthetic_population(series_east_m0, series_west_m0, p_east = 0.2)
)
###
series_east_m65 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTE") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  filter(sex == "male") %>% 
  arrange(year) %>% 
  pull(ex)
series_west_m65 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTW") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  filter(sex == "male") %>% 
  arrange(year) %>% 
  pull(ex)
tmp_df_m65 <- tibble(
  code = "DEUTSYNTH",
  year = 1980:2017,
  x = 65,
  sex = "male", 
  ex = make_synthetic_population(series_east_m65, series_west_m65, p_east = 0.2)
)
hmd_ex_selected_countries_with_synth <- bind_rows(
  hmd_ex_selected_countries,
  tmp_df_m0,
  tmp_df_m65,
  tmp_df_f0,
  tmp_df_f65
)
# Labels for codes 
country_code_lookup <- 
  tribble(
    ~code, ~country,
    "DEUTNP", "Germany",
    "DEUTE", "East Germany",
    "DEUTW", "West Germany",
    "ESP", "Spain",
    "FRATNP", "France", 
    "ITA", "Italy",
    "GBRTENW", "England & Wales",
    "GBR_SCO", "Scotland",
    "DEUTSYNTH", "Synthetic Germany"
  )
```

@@

```{r, echo = FALSE, fig.cap = "Life Expectancy at birth"}
hmd_ex_selected_countries_with_synth %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy"))) %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 0) %>% 
  ggplot(aes(x = year, y = ex, group = sex, colour = sex, linetype = sex)) + 
  geom_line() +
  facet_wrap(~country) +
  labs(
    x = "Year",
    y = "Life Expectancy at birth",
    title = "Life expectancies at birth for selected nations",
    subtitle = "1980 to 2020 or latest available year",
    caption = "Source: Human Mortality Database"
  )
```

```{r, echo = FALSE, fig.cap = "Life Expectancy at age 65"}
hmd_ex_selected_countries_with_synth %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy"))) %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  ggplot(aes(x = year, y = ex, group = sex, colour = sex, linetype = sex)) + 
  geom_line() +
  facet_wrap(~country) +
  labs(
    x = "Year",
    y = "Life Expectancy at age 65",
    title = "Life expectancies at age 65 for selected nations",
    subtitle = "1980 to 2020 or latest available year",
    caption = "Source: Human Mortality Database"
  )
```

@@

```{r, echo = FALSE, warning=FALSE, message = FALSE, fig.height=8, fig.width=8}
change_in_ex_selected_countries %>% 
  filter(x == 0) %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy"))) %>% 
  filter(between(year, 1980, 2020)) %>% 
  ggplot(aes(x = year, y = delta_ex, colour=sex)) + 
  geom_point(alpha = 0.2) + 
  stat_smooth() + 
  facet_grid(country~sex) + 
  geom_hline(yintercept = 0) + 
  theme(strip.text.y = element_text(angle = 0)) + 
  labs(
    x = "Year",
    y = "Year-on-year change in life expectancy",
    title = "Annual change in life expectancy at birth, selected countries",
    subtitle = "Blue line: nonlinear smoother",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )
```

@@

```{r, echo = FALSE, warning=FALSE, message = FALSE, fig.height=8, fig.width=8}
change_in_ex_selected_countries %>% 
  filter(x == 65) %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy"))) %>% 
  filter(between(year, 1980, 2020)) %>% 
  ggplot(aes(x = year, y = delta_ex, colour=sex)) + 
  geom_point(alpha = 0.2) + 
  stat_smooth() + 
  facet_grid(country~sex) + 
  geom_hline(yintercept = 0) + 
  theme(strip.text.y = element_text(angle = 0)) + 
  labs(
    x = "Year",
    y = "Year-on-year change in life expectancy",
    title = "Annual change in life expectancy at age 65, selected countries",
    subtitle = "Blue line: nonlinear smoother",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )
```

@@

```{r, echo = FALSE, warning = FALSE, message = FALSE}
hmd_ex_selected_countries_with_synth %>% 
    filter(year >= 1979) %>% 
    filter(!(code %in% c("DEUTE", "DEUTW", "DEUTNP"))) %>%  # Using only synthetic germany for longer time period
    group_by(code, x, sex) %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
    nest() %>% 
    mutate(
      rwd_model = map(data, ~lm(delta_ex ~ 1, data = . )),
      ar_model  = map(data, ~lm(delta_ex ~ lag(delta_ex), data = .))
    ) %>% 
    mutate(
      aic_rwd = map_dbl(rwd_model, AIC),
      aic_ar  = map_dbl(ar_model, AIC)
    ) %>% 
    mutate(
      diff_aic = aic_ar - aic_rwd
    )  %>% 
  # filter(diff_aic > 0)
  mutate(
    tidied_ar_model = map(ar_model, broom::tidy)
  ) %>% 
  unnest(tidied_ar_model) %>% 
  filter(term == "(Intercept)") %>% 
  arrange(desc(estimate)) %>%
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy"))) %>% 
  select(code, country, x, sex, estimate, std.error) %>% 
  mutate(display_label = glue::glue("{country}_{x}_{sex}")) %>% 
  ggplot(aes(estimate, fct_reorder(display_label, estimate))) + 
  geom_point(aes(colour = x), show.legend = FALSE) +
  expand_limits( x = 0) + 
  geom_vline(xintercept = 0) + 
  labs(
    x = "Average annual life expectancy change",
    y = "Country-sex combination",
    title = "Average annual improvements by country and sex",
    subtitle = "Range: 1980 to 2020 or nearest available years",
    caption = "Source: Human Mortality Database"
  ) + 
  facet_wrap(~x)

@@

@@

NB: those with statistically significant 'breakpoints' in annual change in life expectancy are listed in the table above.

@@

  ex_SCO + labs(x = "Year", y = "Annual change in life expectancy", title = "Scotland", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
```

@@

  ex_FRA + labs(x = "Year", y = "Annual change in life expectancy", title = "France", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
```

@@

  ex_ESP + labs(x = "Year", y = "Annual change in life expectancy", title = "Spain", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
```

@@

  ex_ITA + labs(x = "Year", y = "Annual change in life expectancy", title = "Italy", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
```

@@

 #Grid
 ex_DEUTSYNTH <- hmd_ex_selected_countries_with_synth %>%
      filter(sex %in% c("female", "male")) %>%
      filter(code == "DEUTSYNTH") %>%
    group_by(sex, x) %>% arrange(year) %>%
      mutate(delta_ex = ex - lag(ex)) %>%
    ungroup %>%
    filter(year >= 1979) %>%
    ggplot(aes(x = year, y = delta_ex)) + geom_point() +
   geom_line() +
   geom_hline(yintercept = 0) +
    stat_smooth(mapping = aes(colour = sex)) +
       facet_grid(x ~ sex)

  ex_DEUTSYNTH + labs(x = "Year", y = "Annual change in life expectancy", title = "Germany: Synthetic", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
```

@@

  ex_DEUTE + labs(x = "Year", y = "Annual change in life expectancy", title = "Germany: East", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
```

@@

  ex_DEUTW + labs(x = "Year", y = "Annual change in life expectancy", title = "Germany: West", subtitle = "Life expectancy changes at birth (0) and aged 65 (65) by sex", caption = "Source: Human Mortality Database")
```

## Average improvements by decade

```{r, echo=FALSE, tab.cap = "Source: HMD"}
hmd_ex_selected_countries_with_synth %>% 
    filter(year >= 1979) %>% 
    group_by(code, x, sex) %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>%
    mutate(decade = year - (year %% 10)) %>% 
    filter(!is.na(delta_ex)) %>% 
  group_by(code, x, sex, decade) %>% 
  summarise(
    mean_delta_ex = mean(delta_ex),
    sd_delta_ex   = sd(delta_ex)
    ) %>% 
  ungroup() %>% 
  mutate(cell_contents = glue::glue("{round(mean_delta_ex, 3)} ({round(sd_delta_ex, 3)})")
  ) %>% 
  filter(code == "GBRTENW") %>% 
  filter(x == 0) %>% 
  select(sex, decade, cell_contents) %>% 
  spread(decade, cell_contents) %>% 
#  write.csv("clipboard")
  flextable() %>%
  # Want to make the header row bold
  vline(j = 1, part = "all") %>%
  set_caption(caption = "Average improvement in life expectancy by decade")

```{r, cache=TRUE}
dta_for_decade_comparison <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 0) %>% 
      filter(sex == "female") %>% 
      filter(code == "GBRTENW") %>% 
      filter(year >= 1979) %>% 
      # ggplot(aes(x = year, y = ex)) +
      #   geom_line()
      arrange(year) %>% 
      mutate(delta_ex = ex - lag(ex)) %>% 
    #   ggplot(aes(x = year, y = delta_ex)) + 
    # geom_point() + geom_line() +
    # geom_hline(yintercept = 0)
      mutate(
        decade = year - (year %% 10)
      ) %>% 
  filter(!is.na(delta_ex))
model_null <- 
  lm(delta_ex ~ 1, data = dta_for_decade_comparison)
model_alt <- 
  lm(delta_ex ~ factor(decade), data= dta_for_decade_comparison)
summary(model_null)
summary(model_alt)
anova(model_null, model_alt)
# to get the value of interest do tmp[2,6]
# Generalising 
tidy_dataset <- function(df){
  df %>% 
      filter(year >= 1979) %>% 
      arrange(year) %>% 
      mutate(delta_ex = ex - lag(ex)) %>% 
      mutate(
        decade = year - (year %% 10)
      ) %>% 
  filter(!is.na(delta_ex))
}
make_null_model <- function(df){
  lm(delta_ex ~ 1, data = df)
}
all_null_alt_models <-
  hmd_ex_selected_countries_with_synth %>%
    group_by(x, sex, code) %>% 
    nest() %>% 
    mutate(
      prep_data = map(data, tidy_dataset) 
    ) %>% 
    mutate(
      model_null = map(prep_data, ~lm(delta_ex ~ 1, data = .)),
      model_alt  = map(prep_data, ~lm(delta_ex ~ factor(decade), data = .))
  #    model_null = map(prep_data, make_null_model)
    ) %>% 
    mutate(
      anova_results = map2(model_null, model_alt, anova)
    ) %>% 
    mutate(
      anova_pval = map_dbl(anova_results, ~.[2,6])
    ) 
all_null_alt_models %>% 
  filter(anova_pval < 0.05) -> tmp
summary(tmp$model_alt[[1]])
summary(tmp$model_alt[[2]])
  # build null model
  # build alt model
  # do anova
  # get p value from anova

```{r}
# Let's start with a single example population. (There are many different population subgroups, hence the 
# benefit of being able to automate this later.) We're going to start with England & Wales, females, life 
# expectancy at birth (x == 0)
example_df <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 0) %>% 
      filter(sex == "female") %>% 
      filter(code == "GBRTENW") %>% 
      filter(year >= 1979)
# Now we create the change in ex variable as that's the response variable in the model
example_df <- 
  example_df %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) # %>% 
  # ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
  # geom_line() +
  # geom_hline(yintercept = 0) + 
  # stat_smooth()
# The null model is really simple: just an intercept 
model_null <- lm(delta_ex ~ 1, data = example_df)
summary(model_null)
# 
# The code below shows how we can be even simpler, and force the regression model not to 
# even have an intercept (i.e. have zero predictor variables)
# model_nuller <- lm(delta_ex ~ 0, data = example_df)
# summary(model_nuller)
# And we can compare the intercept with the no intercept model as follows 
# anova(model_nuller, model_null)
# a statistically significant p value shows the intercept is definitely worth including in the 
# model. Substantively that it's wrong to assume no improvement on average in ex over time since the 1980s
# There are lots of possible alternative models, some examples of which are below:
model_alt01 <- lm(delta_ex ~ year >= 1990, data = example_df)
summary(model_alt01)
model_alt02 <- lm(delta_ex ~ year >= 1995, data = example_df)
summary(model_alt02)
model_alt03 <- lm(delta_ex ~ year >= 1997, data = example_df)
summary(model_alt03)
model_alt04 <- lm(delta_ex ~ year >= 2008, data = example_df)
summary(model_alt04)
model_alt05 <- lm(delta_ex ~ year >= 2011, data = example_df)
summary(model_alt05)
# each of these models allows an adjustment to the horizontal line after a given year has been reached. 
# In the examples above some possible given years are 1990, 1995, 1997, 2008 and 2011. 
# Each of these is a candidate value for tau. But we've not looked at all possible/plausible values of this. 
# Hence, it would be good to automate further to search for each possible tau and see how the fit and model 
# results change. 
# We'll start to automate by building a function which creates T_param given tau, and runs and returns the model
# with this type of T_param
# (If tau is 1990, then T_param is equivalent to model_alt01 above; if tau were 1995, then it's equivalent to model_alt02, and so on. The only difference is that the new coefficient created has the same name, T_param, 
# whereas in the above it would have a different name each time, because the coefficient is being 
# created in situ, within the formula specification, rather than by pre-processing the dataframe)
run_alt_with_given_tau <- function(tau, df){
  df <- 
    df %>% 
      filter(!is.na(delta_ex)) %>% 
      mutate(
        T_param = ifelse(year < tau, FALSE, TRUE)
      )

  lm(delta_ex ~ T_param, data  = df)
}
# We can now run the above function, which creates a model, for a wide range of possible tau values
# This is the sequence listed and named tau_candidate, which is every whole number from 
# 1982 to 2015
# For each of these models, we can then get a measure of model fit. We're going to use AIC, though 
# could have used other measures. With AIC (and with BIC, and RMSE etc) better fit is indicated with 
# a lower value
grid_search_df <- tibble(
  tau_candidate = 1982:2015 # column of candidate values
) %>% 
  mutate(
    model_alt = map(tau_candidate, run_alt_with_given_tau, df = example_df) # run the model with the value of tau in tau_candidate
